Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-15 paper/sheaf_lics.tex
section 2 completed

  Sheafification is a popular tool in topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been developed in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher topoi.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, without axiom, being
  compatible with the full type-theoretic axiom of choice.

@@

In other technical difficulty is the fact that sheafification in topos
theory uses the fact that epimorphisms are coequalizers of their
kernel pairs. This properties does not holds in higher topoi, but it
can be replaced by Giraud's axiom which basically allows to replace 
 kernel pairs by \v{C}ech nerves in the property.

@@

Section~\ref{sec:hott} defines the foundations of homotopy type theory
and presents a particular case of GIraud's axiom. 
%
Section\ref{sec:lexmod} introduces the notion of left-exact stratified
modalities, on which this paper is based.
%
Section~\ref{sec:sheaves} defines Lawvere-Tierney sheaves in homotopy
type theory.
%
Section~\ref{sec:sheafification} explains the construction of the
sheafification functor.
%
Section~\ref{sec:formalization} discusses the details of the
formalization in Coq, in particular issues concerning universe
polymorphism.
%
Finally, Section~\ref{sec:future-works} concludes on future works.

@@

In this section, we review some basic definitions in homotopy type
theory that are central in our formalization but not specific to
sheafification. 
%
Section~\ref{ssec:hott} introduces the notion of homotopy types and
classifying objects as defined in~\cite{sets_in_hott}.
%
Section~\ref{sec:colim-homot-type} introduces the notion of colimits
in homotopy type theory. 
%
And Section~\ref{sec:giraud-ax} defines (a particular case of)
Giraud's axiom on higher topoi that we need in the definition of
sheafification.
%
The definition of Section~\ref{ssec:hott} are part of the Coq library
available at \url{https://github.com/HoTT/HoTT}, while other definitions
and theorems are specific to our formalization.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}(b)$ of a function $f$ at element $b$ is
defined as 
$$
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
$$ 
%
A function $f$ is with $n$-truncated homotopy fibers when $\fib{f}{b}$
is an $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
$$\left(\sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n~\fib{f}{b}\right) \simeq (B \to \Type_n)$$
%
and a pullback (with $\Type_n^\bullet \defeq \sum_{A:\Type_n} A$ the universe of pointed
$n$-truncated types)~:
$$
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
$$
for any $f$ with $n$-truncated homotopy fibers, with 
$$t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).$$

@@

In the definition of (a special case of) Giraud's axiom given in
Section~\ref{sec:giraud-ax}, the colimit of the {\em \v{C}ech nerve}
plays a central role.
% 
Following the notion of (homotopy) limits of diagrams over graphs that
has been defined in homotopy type theory~\cite{lumsdaine}, we present
here a definition of colimits of diagrams overs graphs. 
%
The main difference between limits and colimits is that limits are
simply given by $\Sigma$-types, and thus exists already in traditional
type theory, whereas the situation is more complicated for colimits as
it requires the use of higher inductive types.

@@

An colimit of a diagram $D$ over a graph $G$ is given by a type $P$
that defines a cone of $D$, plus the universal property that for any
type $X$, the canonical map that transforms a function from $f : P
\rightarrow X$ to a cone of $D$ on $X$ is an isomorphism.
% 
\begin{defi}
Let $G$ be a graph, and $D$ be a diagram on $G$. 
Let $P:\Type$ together with
\begin{itemize}
\item a map $q_i : D_0(i) \to P$ for any
vertex $i:G_0$, \ie $$q : \prod_{i:G_0} D_0(i) \to P$$
\item for any vertices $i,j:G_0$ and all edges $\phi:G_1(i,j)$, a path
  $p_{i,j}^\phi : q_j \circ D_1(\phi) = q_i$, \ie
  $$p : \prod_{i,j:G_0} \prod_{\phi:G_1(i,j)} q_j \circ D_1(\phi) = q_i.$$
\end{itemize}

@@

Using higer inductive types, every diagram $D$ on a graph $G$ admits a
colimit in homotopy type thoery. This colimit $\colim D$ is given
%
\begin{itemize}
\item a function $q:\prod_{i:G_0} D_0(i) \to \colim D$
\item for any $i,j:G_0$ and $\phi:G_1(i,j)$, a path
  $q_j \circ D_1(\phi) = q_i$.
\end{itemize}
%
Although we have proven in our formalization that $\colim D$ is
actually a colimit over $D$, we do not detail the proof as the
existence of colimits is not use in the definition of the
sheafification process.

\subsection{Giraud's axiom}
\label{sec:giraud-ax}

\begin{defi}
  Let $f:X \to Y$ be a map, and $p:\nat$. The $p$-pullback of $f$,
  noted $X\times_Y \cdots\times_Y X$ is
  the limit of the diagram
  $$\xymatrix{
    X \ar[rd]^f & \cdots & X \ar[ld]_f \\
      &    Y   &
  }$$
  with $p$ copies of $X$. The $0$-pullback of $f$ is $\Unit$.
\end{defi}

In homotopy type theory, we have 
$$X\times_Y \cdots\times_Y X = \sum_{x:X^n} (f x_1 = f x_2) \land
\cdots \land (f x_{n-1} = f x_n).$$
\begin{defi}
  If $f:X \to Y$ is a map, then the {\em \v{C}ech nerve} of $f$ is the
  diagram
  $$\xymatrix{
    C(f) := \cdots~ X \times_Y X \times_Y X \tar[r]{4pt} & X \times_Y X \dar[r]{2pt} & X
  }$$
where the arrows are the canonical projections.
\end{defi}

\begin{ax}[Giraud]
  If $f:X\to Y$ is a surjection, then the
colimit of $C(f)$ is $Y$.
\end{ax}

Viewing homotopy type theory as a
type-theoristic version of higher topos theory, \cite{lurie} suggests that this
axiom can be proved, although we admit it in this paper.

>>>>

  Sheafification is a popular tool in topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been developed in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher topoi.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, being
  compatible with the full type-theoretic axiom of choice.

@@

In other technical difficulty is the fact that sheafification in topos
theory uses the fact that epimorphisms are coequalizers of their
kernel pairs. This properties does not holds in higher topoi, but
using Giraud-Rezk-Lurie axioms of $\infty$-topoi, the same property
holds by replacing kernel pairs by \v{C}ech nerves in the statement.

@@

Section~\ref{sec:hott} defines the foundations of homotopy type theory
and presents a particular case of Giraud-Rezk-Lurie axioms. 
%
Section~\ref{sec:lexmod} introduces the notion of left-exact stratified
modalities, on which this paper is based.
%
Section~\ref{sec:sheaves} defines Lawvere-Tierney sheaves in homotopy
type theory.
%
Section~\ref{sec:sheafification} explains the construction of the
sheafification functor.
%
Section~\ref{sec:formalization} discusses the details of the
formalization in Coq, in particular issues concerning universe
polymorphism.
%
Finally, Section~\ref{sec:future-works} concludes on future works.

@@

In this section, we review some basic definitions in homotopy type
theory that are central in our formalization but not specific to
sheafification. 
%
Section~\ref{ssec:hott} introduces the notion of homotopy types and
classifying objects as defined in~\cite{sets_in_hott}.
%
Section~\ref{sec:colim-homot-type} introduces the notion of colimits
in homotopy type theory. 
%
And Section~\ref{sec:giraud-ax} defines a particular case of
Giraud-Rezk-Lurie axioms on higher topoi that we need in the
definition of sheafification.
%
The definition of Section~\ref{ssec:hott} are part of the Coq library
available at \url{https://github.com/HoTT/HoTT}, while other definitions
and theorems are specific to our formalization.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
$$
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
$$ 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is an $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
$$\chi : \left(\sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType \
\fib{f}{b}\right) \xrightarrow{\sim} 
 (B \to \Type_n)$$
%
 such that the following diagram is a pullback for any $f$ with
 $n$-truncated homotopy fibers:
$$
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
$$
with $\Type_n^\bullet \defeq \sum_{A:\Type_n} A$ the universe of pointed
$n$-truncated types and 
$$t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).$$

@@

In the definition of (a special case of) Giraud-Rezk-Lurie axiom given in
Section~\ref{sec:giraud-ax}, the colimit of the {\em \v{C}ech nerve}
plays a central role.
% 
Following the notion of (homotopy) limits of diagrams over graphs that
has been defined in homotopy type theory~\cite{lumsdaine}, we present
here a definition of colimits of diagrams overs graphs. 
%
The main difference between limits and colimits is that limits are
simply given by $\Sigma$-types, and thus exists already in traditional
type theory, whereas the situation is more complicated for colimits as
it requires the use of higher inductive types.

@@

A colimit of a diagram $D$ over a graph $G$ is given by a type $P$
that defines a cone of $D$, plus the universal property that for any
type $X$, the canonical map that transforms a function $f : P
\rightarrow X$ to a cone of $D$ on $X$ is an isomorphism.
% 
\begin{defi}
Let $G$ be a graph, and $D$ be a diagram on $G$. 
Let $P:\Type$ together with
\begin{itemize}
\item a map $q_i : D_0(i) \to P$ for any
vertex $i:G_0$, \ie $$q : \prod_{i:G_0} D_0(i) \to P$$
\item for any vertices $i,j:G_0$ and all edges $\phi:G_1(i,j)$, a path
  $p_{i,j}^\phi : q_j \circ D_1(\phi) = q_i$, \ie
  $$p : \prod_{i,j:G_0} \prod_{\phi:G_1(i,j)} q_j \circ D_1(\phi) = q_i.$$
\end{itemize}

@@

Using higer inductive types, every diagram $D$ on a graph $G$ admits a
colimit in homotopy type thoery. This colimit $\colim D$ is given
%
\begin{itemize}
\item a function $q:\prod_{i:G_0} D_0(i) \to \colim D$
\item for any $i,j:G_0$ and $\phi:G_1(i,j)$, a path
  $q_j \circ D_1(\phi) = q_i$.
\end{itemize}
%
Although we have proven in our formalization that $\colim D$ is
actually a colimit over $D$, we do not detail the proof here as the
existence of colimits is not use in the definition of the
sheafification process. We only make use of the following special case
of Giraud-Rezk-Lurie axioms.

\subsection{On Giraud-Rezk-Lurie axioms}
\label{sec:giraud-ax}

The Giraud-Rezk-Lurie axioms are the $\infty$-version of Giraud's
axioms that characterize a topos. Namely, there are 4 axioms on a
$(\infty,1)$-category that have been shown to be equivalent to be
equivalent to $(\infty,1)$-topos~\cite[Chapter 6]{lurie}. This is not
the purpose of this paper to discuss those axioms, we only state here
a particular case of the last axiom which says that every groupoid
object is effective. 
%
This particular case connects an epimorphism (or $-1$-truncated
function) to the colimit of its \v{C}ech nerve. 
%
This is the $\infty$-generalization of the fact that, in a topos,
every epimorphism is the colimit of its kernel pair, which plays a
central part in the definition of sheafification.

\begin{defi}
  Let $f:X \to Y$ be a map, and $p:\nat$. The $p$-pullback of $f$
  (also called the $p$-fold fiber product of $X$ over $Y$),
  noted $X\times_Y \cdots\times_Y X$, is
  the limit of the diagram
  $$\xymatrix{
    X \ar[rd]^f & \cdots & X \ar[ld]_f \\
      &    Y   &
  }$$
  with $p$ copies of $X$. The $0$-pullback of $f$ is $\Unit$.
\end{defi}

As explained in Section~\ref{sec:colim-homot-type}, every limit exists
in (homotopy) type theory and can be described by a $\Sigma$-type. 
%
The particular case of a $p$-pullback is given by
%
$$
X\times_Y \cdots\times_Y X = \sum_{x:X^n} (f x_1 = f x_2) \land
\cdots \land (f x_{n-1} = f x_n).
$$
%
The \v{C}ech nerve of a map $f$ can then be described as the
simplicial object that, at degree $p$, is given by the
$(p+1)$-pullback of $f$. The \v{C}ech nerve can thus also be described
as a diagram over $\Delta^{op}$ (where $\Delta$ is the simplicial
category, seen as a graph). For simplicity, we only give an intuitive
definition here, the complete definition can be found in our
formalization.
%
\begin{defi}
  Let $f$ be a map from $X$ to $Y$. The {\em \v{C}ech nerve} $C(f)$ of $f$
  is given by the diagram
  $$\xymatrix{
    C(f) := \cdots~ X \times_Y X \times_Y X \tar[r]{4pt} & X \times_Y X \dar[r]{2pt} & X
  }$$
where the arrows are the canonical projections.
\end{defi}

\begin{ax}[Giraud-Rezk-Lurie]
  For any epimorphism $f : X \to Y$, the colimit of its \v{C}ech nerve
  $C(f)$ is $Y$.
\end{ax}

With the point of view that homotopy type theory is a type-theoristic
version of higher topos theory, Theorem 6.1.0.6 of~\cite{lurie}
suggests that this axiom can be proved in homotopy type theory. 
%
But proving this axiom seems to be a quite complex task on its own,
and we have decided to admit it in this paper.
